---
title: "Custom Databases: Classification & Analyses"
author: "Bastiaan"
date: "27 March 2021"
output: html_document
runtime: shiny
---

Introduction
------------

The following R markdown file performs the steps needed to:

1. Obtain taxonomic information and Naturalis specimen records for selected taxa,
2. Create a data structure capable of providing highly visual representations
and at the same time maintaining integrity of each data set’s origin,
3. Search out and visualize the overlap / discrepancies between obtained public
reference data and records readily available at Naturalis Biodiversity Center.

The following libraries need to be loaded:

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
library(rmarkdown)
library(data.table)
library(taxizedb)
library(myTAI)
library(tidyr)
library(shiny)
library(DT)
library(plyr)
library(dplyr)
library(stringr)
library(d3Tree)
library(billboarder)
library(nbaR)
```

In addition, here We define the global variables that we will reuse throughout
the code:

```{r globals, include=FALSE}
# This should define the root folder of the local copy of the git repository at:
# https://github.com/naturalis/Custom-databases-DNA-sequences, which is
# automatically defined correctly if we run the present code from within a
# local clone of the repo and have set the working directory to the script
# source (in RStudio: Session > Set working directory > To source file location)
setwd('..')
REPO_HOME <- paste(dirname(getwd()), sep = "")

getwd()
```

### Load data

We now load the following data sets:

1. Reference Data Set (NSR),
2. Obtained Public specimen records and sequence data (BOLD),
3. List of Marker Codes,
4. List of Public Databases

Both the reference data set (1) and the obtained public reference data (2) are
output from the [script](../script/custom_databases.py)

```{r datasets, include=FALSE}
## Load species and synonyms from the NSR files, respectively

nsrSynonyms.file <- sprintf('%s/data/insert_files/nsr_synonym.csv', REPO_HOME)
nsrSynonyms <- read.csv(nsrSynonyms.file, encoding="UTF-8")
nsrSpecies.file <- sprintf("%s/data/insert_files/nsr_species.csv", REPO_HOME)
nsrSpecies <- read.csv(nsrSpecies.file, encoding="UTF-8")


```

### NSR Backbone

To facilitate taxonomic estimation by lowest common ancestor approach, the 
complete classification of species should ideally be present in a reference 
taxonomic data set. As the NSR export only held the scientific names of all
species of interest, the Netherlands Biodiversity API will be queried to obtain
their higher classification.

```{r nsrBackbone, include=FALSE}
## Removed the last empty row by hand (error)
nsrTaxonomy.file <- sprintf('%s/data/input_files/NSR_taxonomy1.csv', REPO_HOME)
## Skip first few lines untill header
nsrTaxonomy <- read.csv(nsrTaxonomy.file, encoding="UTF-8", skip = 4, sep = "",  na.strings="NA")
## Get the necessary rows (scientific_name, common_name, rank, nsr_id)
nsrTaxonomy <- subset(nsrTaxonomy, select = c(1,2,4,6))  
nsrTaxonomy <- nsrTaxonomy %>% filter(rank == "soort")

## Split genus, species and identification reference
nieuw <- str_split_fixed(nsrTaxonomy$scientific_name., " ", 3)
## Drop empty lines (CHECK)
df_mtx <- as.data.frame(nieuw) %>% drop_na()
## Give columnnames
colnames(df_mtx) <- c("genus", "species", "identification_reference")
 
## Remove parentheses for every identification reference 
df_mtx$identification_reference <- gsub('[()]','',df_mtx$identification_reference) 

## Merge the genus and species to binomial in new column
df_mtx$species_name <- paste(df_mtx$genus, df_mtx$species)

## Keep rows where species do not dots or parenthesis in their name
df_mtx <-  df_mtx[grep('[()\\.]', df_mtx$species, invert = TRUE),]

## Remove the × symbol in species_name from all the rows (are there for three species names)
df_mtx$species_name <- gsub('[\\×]','',df_mtx$species_name)

## Keep rows where identification references do not have [] in their name (for an instance with [unknown])
df_mtx <-  df_mtx[grep('[\\[\\]', df_mtx$identification_reference, invert = TRUE),]

## Fill empty strings with NA's so these rows can be dropped
df_mtx <- df_mtx  %>% mutate_all(na_if,"")

## Drop rows containing NA's
x <- na.omit(df_mtx)

## Add unique identifier counting from 0
x$species_id <- seq.int(from = 0, to = nrow(x)-1)

## Keep and rearrange columns (species_id, species_name, identification_reference)
x <- subset(x, select = c(5, 4, 3))  


## Temp: Check what the differences are from B. Ankers to this one.
difference <- (!nsrSpecies$species_name %in% x$species_name)
not.in.x <- nsrSpecies[difference,]
difference <- (!x$species_name %in% nsrSpecies$species_name)
not.in.nsr <- x[difference,]



nsrSpecies[,"species_name"] # can now be x[,"species_name"]
## Select species for taxon data retrieval
nsrIn <- tolower(nsrSpecies[, "species_name"])
head(nsrIn, 3)

## Query NSR taxon records (duration: ~40m)
## Method: nbaR 'taxon_query'
## Parameters: NSR source code and list of selected species
queryParams <- list("sourceSystem.code"="NSR")
nsrSpecimens <- data.frame()
for(x in 1:length(nsrIn)){
  queryParams[["acceptedName.scientificNameGroup"]] <- nsrIn[x]
  nsrSpecimens <- bind_rows(nsrSpecimens, taxon_query(queryParams))
}

## Access and extract required taxonomic fields from retrieved taxon records
nsrBackbone = data.frame()
for(x in 1:nrow(nsrSpecimens)){
  tryCatch({
    c.id=nsrSpecimens$sourceSystemId[x]
    c.kingdom=nsrSpecimens$defaultClassification$kingdom[x]
    c.phylum=nsrSpecimens$defaultClassification$phylum[x]
    c.class=nsrSpecimens$defaultClassification$className[x]
    c.order=nsrSpecimens$defaultClassification$order[x]
    c.family=nsrSpecimens$defaultClassification$family[x]
    c.genus=nsrSpecimens$defaultClassification$genus[x]
    c.species=nsrSpecimens$acceptedName$specificEpithet[x]
    c.authority=nsrSpecimens$acceptedName$authorshipVerbatim[x]

    row <- data.frame(cbind(tax_id=c.id,kingdom=c.kingdom,phylum=c.phylum,class=c.class,
                        order=c.order,family=c.family,genus=c.genus,
                        species=paste(c.genus,c.species),identification_reference=c.authority))
    nsrBackbone <- bind_rows(nsrBackbone, row)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
## Remove parenthesis around the authority for consistency across data sets
nsrBackbone$identification_reference = gsub("[()]", "", nsrBackbone$identification_reference)

head(nsrBackbone,3)

## Match taxonomic data to respective species ID, by species name
nsrBackbone <- left_join(nsrBackbone, nsrSpecies[1:2], by=c("species"="species_name")) %>%
  .[,c(1,10,2,3,4,5,6,7,8,9)]
nsrBackbone <- unique(nsrBackbone)
## nsrBackbone after binding them to their species id by name (nsrSpecies), 10 columns (tax_id got added)

## Write NSR data to file
write.csv(nsrBackbone, file=sprintf('%s/data/insert_files/tree_nsr.csv', REPO_HOME),
          row.names=FALSE, fileEncoding="utf-8")

## Clean up query variables
rm(x, row, queryParams, list = ls()[grep("^c.", ls())])

```

### Naturalis Coverage

To determine the overlap / discrepancies between the obtained public sequence data
and the reference at Naturalis Biodiversity Center, in relation to the NSR, we now
query the Naturalis specimen records (using the Netherlands Biodiversity API).

```{r natCoverage, include=FALSE}
## Select species for taxon data retrieval
## Ensuring to select only binomial names
natIn <- unique(gsub("([A-Za-z]+).*", "\\1", nsrSpecies$species_name))
## Query Naturalis specimen records (duration: ~20m)
## Method: nbaR 'specimen_query'
## Parameters: CRS source code and list of selected species

queryParams <- list("sourceSystem.code"="CRS") #removed queryParams <- list(queryparams) before <- list(sourceS.. etc)
natSpecimens <- data.frame()
for(x in 1:length(natIn)){
  queryParams[["identifications.defaultClassification.genus"]] <- natIn[x]
  specimen_query(queryParams)
  natSpecimens <- bind_rows(natSpecimens, specimen_query(queryParams))
}

## natSpecimens (13 rows, with @CRS!!)

## Access and extract required data fields from retrieved specimen records
natSpecies = data.frame()
for(x in 1:nrow(natSpecimens)){
  tryCatch({
    c.genus=natSpecimens$identifications[[x]]$defaultClassification$genus
    c.species=natSpecimens$identifications[[x]]$defaultClassification$specificEpithet
    c.count=natSpecimens$numberOfSpecimen[[x]]
    c.id=natSpecimens$sourceSystemId[[x]]
    row <- data.frame(cbind(species=paste(c.genus,c.species),counts=c.count,sequenceID=c.id))
    natSpecies <- bind_rows(natSpecies, row)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
## Remove NAs
natSpecies$counts[is.na(natSpecies$counts)] <- 0
## Update data type and sum unique species record counts
natSpecies$counts <- as.numeric(as.character(natSpecies$counts))
natSpecies <- aggregate(natSpecies$counts,
                        by=list(natSpecies$species,natSpecies$sequenceID),sum)
## Rename columns
colnames(natSpecies) <- c("species", "sequenceID", "counts")
## Adopt respective taxonomic unit for synonymous names
natSpecies <- left_join(natSpecies, nsrSynonyms[2:3], by=c("species"="synonym_name"))
natSpecies <- left_join(natSpecies, nsrSpecies[1:2], by=c("species_id"))
natSpecies$species_name <- coalesce(natSpecies$species_name, natSpecies$species)
natSpecies <- natSpecies[,c(5,2,3)]
## Match species against the NSR
natCoverage <- merge(natSpecies, nsrSpecies[c(2:3)], by="species_name")

write.csv(nsrBackbone, file=sprintf('%s/data/exports/naturalis.csv', REPO_HOME),
          row.names=FALSE, fileEncoding="utf-8")

## Clean up query variables
rm(x, row, queryParams, list = ls()[grep("^c.", ls())])
```

### Additional taxonomy

The ability to divert to a different backbone provides additional insight into a
specimens classification or serves as a filter for specific taxa (e.g., marine: WoRMS).
Additional taxonomic hierarchies can be retrieved from various databases.

```{r tree_ncbi, include=FALSE}
## Download taxonomic database (options: ncbi, itis, gbif, col, wfo)
## Used data source: NCBI
db_download_ncbi()

## Load species names from the NSR data set
taxidIn <- nsrSpecies[, "species_name"]

## Match species names to NCBI taxon IDs
taxidOut <- taxizedb::name2taxid(taxidIn, db="ncbi", out_type="summary")
head(taxidOut,3)
## NCBI taxon ID's

## Isolate taxon IDs from output, save as vector
treeIn <- as.vector(taxidOut$id)
head(treeIn,3)
## Only id's

## Retrieve taxonomic hierarchy for each taxon ID
treeOut <- taxizedb::classification(treeIn, db="ncbi", row=1, verbose=FALSE)
head(treeOut,3)
## Parse out the taxonomy levels/factors that you require
taxdata = data.frame()
for(x in 1:length(treeOut)){
  tryCatch({
    for(i in 1:length(treeOut[[x]][[1]])) {
      c.tax_id=filter(treeOut[[x]])$id[[i]]
      c.parent_tax_id=NULL
      c.rank=filter(treeOut[[x]])$rank[[i]]
      c.name=filter(treeOut[[x]])$name[[i]]

      tryCatch({
        c.parent_tax_id=filter(treeOut[[x]])$id[[i-1]]
        }, error=function(e){c.parent_tax_id=NULL})
      
      row <- data.frame(cbind(tax_id=c.tax_id,parent_tax_id=c.parent_tax_id,
                            rank=c.rank, name=c.name))

      # Check if record exists
      if (nrow(merge(row,taxdata))==0) {
        taxdata <- bind_rows(taxdata, row)
      }
    }
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

## Match hierarchies to NSR species names by taxon IDs
## (Accounting for use of synonymous names within NCBI)
ncbiBackbone <- left_join(taxdata, taxidOut, by=c("tax_id"="id"),
                       suffix=c("","_nsr")) %>%
  left_join(., nsrSpecies[1:2], by=c("name_nsr" = "species_name")) %>%
  .[, c(1,6,4,2,3)] # Reaarange columns, drop 'name_nsr' column
ncbiBackbone <- unique(ncbiBackbone)

## Write NCBI data to file
write.csv(ncbiBackbone, file=sprintf('%s/data/insert_files/tree_ncbi.csv', REPO_HOME),
          row.names=FALSE, fileEncoding="utf-8", na="")

## Clean up query variables
rm(x, i, row, list = ls()[grep("^c.", ls())])
```